Visualization and analysis of probability distributions of large temporal data

### 2022-02-27 report/response.Rmd

# Professor Catherine Hurley

## Chapter 1

> This would be a good place to detail the co-authorship arrangements for the succeeding chapters, and for the student’s contribution to be stated clearly.

A table outlining the main chapters’ co-authorship arrangements included in Introduction on page 5.

@@

> Q1. On page 33 there is a mention of the hakear package, without a reference.

No change.
It appears in the abstract of the paper. For the thesis, I have added it since the abstract acts as a summary to the chapter. The package reference is mentioned in the Supplementary section of the paper. Moreover, the plan is to bring all the functions from `gravitas` to `hakear` when we publish it.

> Q2. In the Introduction state the methods are for continuous variables.

Added "for a continuous univariate dependent variable" in the last paragraph of Introduction of Chapter 3 on page 39.

@@

> Q8. Maybe include a reference for gestalt theory.

Done.

> Q9. In Section 3.2.2, I find the references to the null distribution confusing. Surely that is the design in
Figure 3.3(a) only?

Added text in the caption of Figure 3.3(a) and Section 3.2.2 (Notation) on page 41 to make it clear.

<!-- Clarification: Null distributions refer to the case when there is no significant distributional differences between categories of one or more granularities. Data in Fig 3.3 (a) is consistent with the null distribution. -->

@@

Added $m$ to table 3.1  
$m$: number of cyclic granularities to display together

Clarification: $N_c$ refers to the total number of contextual cyclic granularities. $m$ refers to the number of cyclic granularities we are considering together in the display. For example, contextual cyclic granularities could be $hour_{day}$, $day_{week}$ and $month_{year}$ and we want to visualize any one granularity at a time. So $N_c = 3$ and $m=1$.

> Q11. Add v to table 3.1.

Added v:continuous univariate measured variable to the table on page 42.

> Q12. In the first sentence on the top of page 42, ordered and unordered are mixed up. Have you considered the setting where the facet variable levels are ordered?

> For the within-facet ordered distances, you could consider a distance measure that respects circular order, or choose the start level appropriate to the display.

This is a good suggestion. We will include that into future work.

@@

This is a good suggestion. We will include that into future work.

> Q13. I like Figure 3.4. In (3), the dotted arcs only connect to a1 which might be misleading.

Caption changed to address this comment.

<!-- Within-facet distances are illustrated in Panels 3) (when categories are un-ordered, shown only with respect to a1) and Panel 4) (when categories are ordered) -->

@@

> Q14. In the pairwise distance measure on page 43, is there any adjustment made for varying numbers of observations across the levels of A and B? For example in Figure 3.3(d) if there are few values at xlevel=1 the comparison shown is less interesting

Added text in the section 3.2.3 under subsection "Characterizing distributions" on page 43 to address this comment.

<!-- Clarification: As long as that there is enough data for each level or combination of levels, no adjustment for varying number of observations is done. If there are few data for one or more levels, subsequent categories could be collapsed to obtain precise estimates of quantiles.  -->

@@

> Q15. The equations on page 44 could be tidied up. There are extra () on the definition of $wpd_{perm}$ and on the residual definition at the bottom of the page. Use $\times$ for the equations.

Done.

> Q16. Is there a practical reason why $wpd_{glm}$ is not working as expected for lower $nx$ and $nf$?

Added the empirical reason in Section 3.2.4 under sub-section "Combination approach" on page 47.

<!-- Clarification: There is no practical explanation for this, but a more empirical one - with smaller $n_x$ and $n_f$, the variance in the values of $wpd_{raw}$ is greater. As a result, our modeling technique that attempts to fit $wpd_{raw}$ to the number of comparisons ($n_x\times n_f$) performs poorly. The increased volatility in the values of $wpd_{raw}$ with a smaller number of levels is due to the fact that no asymptotic distribution for the statistic "maximum" (which is the foundation of $wpd_{raw}$) across such small number of values could be produced. -->

@@

> Q18. In Table 3.2, I would suggest separating the m = 1 and m = 2 tables.

$m=1$ removed from the main chapter and included in supplements.

> Q19. The presentation of the material on the simulation study on page 48 could be improved. Where is the notation $wpd_{l,s}$ used? Figure 3.5 shows the m = 2 results only. The text alludes to a simulation involving different underlying distributions, but this is not mentioned again.

Rephrased the simulation design and results corresponding to $m=2$ on page 50. Similar design and results for $m=1$, although important, are not included in the main chapter but in the supplements. Simulation results involving different underlying distributions are also presented in the supplementary paper.

@@

> Q4. End of Chapter 4.1: incorrect reference to Section 2.7.

Done.

> Q5. I found Figure 4.1 very useful. In each box, would be helpful to put in “or”, in places where only one of the steps listed is performed, eq Normal quantile transform or Robust scaling. As there are many steps in the algorithm, it would be helpful to the reader to label the pipeline steps and to refer back to them in the text. The text in the Data pre-processing box does not make sense to me.

Rephrased the text in the section 4.2.3 (Data pre-processing) to make it easier to understand. Figure 4.1 labels changed to make it consistent with the text.

@@

> Q6. The section describing RS and NQT is confusing. It states RS is applied to each time series separately. Is NQT also applied to each observation separately? How does “it could be useful to standardize it for the selected set of significant granularities prior to computing the distances” relate to the following bullet points?

Text added under Section 4.2.2 on pages 64 and 65 to clarify this.

<!-- Clarification: Each observation (time series) is scaled independently in both JS-NQT and JS-RS. For the WPD approach, each observation might be scaled or not scaled, resulting in usable groups in either case. -->

@@

> Q7. Page 64 “D is the Jensen-Shannon distance”.

No change. I had mentioned that $D$ stands for Jensen-Shannon distances (now page 66).

@@

> Q8. In Table 4.1, the R column should have 250, not 20.

Done.

> Q9. The description of the data generation on page 67 and Table 4.2 is confusing. Maybe state the distribution of each $v_t$.

The table format and caption changed to address this comment. (page 70)

<!-- No change. Stating the distribution of each $v_t$ is difficult because the levels for each granularity and design varies, therefore I chose to divide it into two tables. -->

@@

> Q10. Where is $\mu$ in Table 4.2?

The changed caption (Table 4.2 page 70) will help clarify this.

@@

> Q11. In Section 4.3.3 what are the values of R and T?

Added in the first line of Section 4.3.3 (page 72.)

> Q12. Figure 4.5 caption: MDS summary plots of what?

Added in the caption (page 75) to address this comment.

<!-- Clarification: MDS summary plots of clusters obtained from employing JS-NQT on the validation data sets. -->

@@

> Q13. Figure 4.7 is missing labels (a), (b) (c).

Added a, b representing 12 customers (rows) in each of them (page 78).

@@

> Q14. Why do you chose to summarize 4 and 5 clusters from JS-NQT when sindex suggests 3 clusters?

Caption (Figure 4.9 page 80) and description in the text (page 78) changed to address this comment.

<!-- Clarification: Fig. 4.9 depicts a visual summary of why three clusters are insufficient for this data set, however it is recommended by sindex summary plots. In a cluster analysis, it is standard practice to test a few cluster sizes before and after the optimal $k$ to see if a different number of clusters is useful. In this example, increasing the number of clusters actually unfolds some behavior which was not apparent with $k=3$ as suggested by sindex. -->

@@

> Q15. The reference to the stationarity assumption in the Discussion needs Clarification.

Text changed in the Discussion section (under Section 4.5.2) to address this.

<!-- Clarification: "we have assumed the time series to be stationary" has been changed to "the conditional distributions are assumed to remain constant for the observation period". -->

@@

> Q16. Page 81 “can not” should be “cannot”

Done.

> Q17. Again, the computational burden could do with more Discussion. At present, if I were to use this method on my data, what kind of sizes are realistic to work with?

A paragraph is added to the Discussion section to incorporate comments on the computational burden. (second paragraph of section 4.5.2 on pages 83, 84)

@@

Done.

> Q5. Before introducing a new distance measure in Section 3.2, I would like to see a literature review of existing distance measures and their limitations. Why is it necessary to introduce a new distance measure here? This is partially addressed on p. 41, but that should be placed earlier in the text. One clarifying question: Which density is represented by f in the equation? And shouldn't there be a dx at the end of the integral? Also, the distance measures that are mentioned at the end of that paragraph should be explained in one or two sentences.

Changed the following:

- Introduction (last but one paragraph on page 39) to incorporate comments on existing distance measures and their limitations

- "Distance between distributions" (under section 3.2.3 on page 44) to mention why we chose we chose Jensen-Shannon distances and some details about other distance measures.

@@

> Q15. The last paragraph on p. 48 uses $<$ while equation (3.2) uses $\leqslant$. Which one is correct?

Done.
$\leqslant$ for both.

> Q16. Match the text on p. 50 and introduce the abbreviations from Figure 3.7 and Table 3.3, e.g., "hod" likely is "hour day" and so on. Introducing these abbreviations on p. 53 comes
too late.

Added the abbreviations on page 54 in "Choosing cyclic granularities of interest and removing clashes" under section 3.4.

@@

Caption rephrased to make it more clear on page 56.

<!-- For id: 7, there is more distributional difference between consecutive hours of the day compared to id:1. Also all the days of the week are different from each other in id:7. id:1 has similar behavior across all days within both weekdays and weekends. For this reason, the wpd for id:7 has a higher value than id:1 as the similar behavior within weekdays and weekends balances out the difference between weekdays and weekends. The display with hod vs wdwnd would have shown higher values of wpd for id:1 (as can be verified from Table 3). -->

> Q22. I didn't ask earlier, but how were the eight households for Section 3.4 being selected? Was this based on a random sample, or manually do obtain rather different energy usage patterns? This should be answered earlier on, but this also should be addressed in the Discussion: What can be expected when working with the full data set of Chapter 2, i.e., about 13,000. You indicate:
A future direction of work is to be able to explore and compare many individuals/subjects together for similar patterns across significant granularities. What would be computationally (and visually) feasible as 13,000 is a few orders of magnitude
higher than the number of eight households used in this example? Clearly, no full answer is expected here, but rather some speculation of what could possibly be done in the future (or what is done in the next chapter).

- Text added in the first paragraph of Application to elaborate how households were selected (page 51).

- A paragraph is added to the Discussion section to incorporate comments on computational time (page 57).

@@

Done.

## Chapter 4

> Q1. There seems to be some contradiction on p. 60. One sentence states: Tureczek and Nielsen (2017) conducted a systematic study of over 2100 peer-reviewed papers on smart meter data analytics. Another sentence states: Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in Tureczek and Nielsen (2017).

Rephrased Section 4.1 on page 62 to make it clear that most techniques mentioned in Tureczek and Nielsen (2017) do not treat smart meter data as a data type with temporal component.

@@

> Q12. It would help the reader to also mention in the caption of Table 4.3 that P-1 & Q-1 and P-2 & Q-2 are identical. I first looked at Figure 4.9 and then at Table 4.3, but if read in the other order, I likely would have been surprised, to see identical values in some of the table rows (without any further explanation).

Done.

> Q13. For the thesis, the work is adequate and meaningful. For a journal paper, I would like to see some of the points outlined in the Discussion being addressed. What happens if this method is applied to the data from the 13,000 customers that were introduced earlier on? As a reader of a journal paper, I would rather like to see limitations of the method proposed here, e.g., if it does not scale up then why it most likely does not scale up. Otherwise, it is hard to assess whether it is worthwhile to try this method for one's own data.

The Discussion section is split into (1) conclusions, and (2) limitations and future research (including a discussion on computational complexity) on pages 82-84 to address this comment.

### 2022-02-27 Rmd/03-hakear.Rmd
final pages added on response document

The simulation results (in Section \ref{sec:simulations}) show that the distribution of $\wpdsub{glm}$ under the null design is similar for high $\nx$ and $\nf$ (levels higher than $5$) but less so for lower values of $\nx$ and $\nf$. Hence, a combination approach is proposed where we use a permutation approach for categories with small numbers of levels, and a modeling approach for categories with higher numbers of levels. This ensures that the computational load of the permutation approach is alleviated while maintaining a similar null distribution across different categories. This approach, however, requires that the adjusted variables from the two approaches are brought to the same scale. We define $\wpdsub{glm-scaled} = \wpdsub{glm}\times \sigma^2_{\text{perm}}/\sigma^2_{\text{glm}}$ as the transformed $\wpdsub{glm}$ with a similar scale as $\wpdsub{perm}$. The adjusted measure from the combination approach, denoted by $\wpd$ is then defined as follows:
\begin{equation}
\wpd = \begin{cases}
         \wpdsub{perm},       & \text{if $\nx, \nf \le 5$};\\
         \wpdsub{glm-scaled}  & \text{otherwise}.\\
       \end{cases}
\end{equation}

>>>>

The simulation results (in Section \ref{sec:simulations}) show that the distribution of $\wpdsub{glm}$ under the null design is similar for high $\nx$ and $\nf$ (levels higher than $5$) but less so for lower values of $\nx$ and $\nf$. An empirical explanation for this is that with smaller $n_x$ and $n_f$, the variance in the values of $wpd_{raw}$ is greater [Figure 13 of supplements](https://github.com/Sayani07/paper-hakear/blob/master/paper/paper-supplementary.pdf). As a result, our modeling approach that fits $wpd_{raw}$ to the number of comparisons ($n_x\times n_f$) performs poorly. In order to address this, a combination approach is proposed where we use a permutation approach for categories with small numbers of levels, and a modeling approach for categories with higher numbers of levels. This ensures that the computational load of the permutation approach is alleviated while maintaining a similar null distribution across different categories. This approach, however, requires that the adjusted variables from the two approaches are brought to the same scale. We define $\wpdsub{glm-scaled} = \wpdsub{glm}\times \sigma^2_{\text{perm}}/\sigma^2_{\text{glm}}$ as the transformed $\wpdsub{glm}$ with a similar scale as $\wpdsub{perm}$. The adjusted measure from the combination approach, denoted by $\wpd$ is then defined as follows:
\begin{equation}
\wpd = \begin{cases}
         \wpdsub{perm},       & \text{if $\nx, \nf \le 5$};\\
         \wpdsub{glm-scaled}  & \text{otherwise}.\\
       \end{cases}
\end{equation}

### 2022-02-27 Rmd/04-gracsr.Rmd
final pages added on response document

The shape and scale of the distribution of the measured variable (e.g. energy usage) affects distance calculations. Skewed distributions need to be symmetrized. Scales of individuals need to be standardized, because clustering is to select similar patterns, not magnitude of usage. (Organizing individuals based on magnitude can be achieved simply by sorting on a statistic like the average value across time.) <!--Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories.--> For the JS-based approaches, two data transformation techniques are recommended, normal-quantile transform (NQT) and robust scaling (RS).<!-- NQT is used in @Gupta2021-hakear prior sorting granularities.-- $\wpd$, which is the foundation of wpd-based distances. XX wpd was not yet introduced--> While @Gupta2021-hakear already use NQT when computing $\wpd$, it could be useful to standardize it for the selected set of significant granularities prior to computing the distances.

@@

All of the approaches were fitted to each data design and to each combination of the considered parameters. The formed clusters have to match the design, be well separated, and have minimal intra-cluster variation. <!--It is possible to study these desired clustering traits visually in a more comprehensive way than just looking at index values.--> MDS and parallel coordinate graphs are used to demonstrate the findings, as well as an index value plot to provide direction on the number of clusters. JS-based approaches corresponding to NQT and RS are referred to as JS-NQT and JS-RS respectively. In the following plots, results for JS-NQT are reported, and results with JS-RS or wpd-based distances are in the supplementary material.

>>>>

The shape and scale of the distribution of the measured variable (e.g. energy usage) affects distance calculations. Skewed distributions need to be symmetrized. Scales of individuals need to be standardized, because clustering is to select similar patterns, not magnitude of usage. (Organizing individuals based on magnitude can be achieved simply by sorting on a statistic like the average value across time.) <!--Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories.--> For the JS-based approaches, two data transformation techniques are recommended, normal-quantile transform (NQT) and robust scaling (RS) referred to as JS-NQT and JS-RS respectively. The approach based on $\wpd$ is referred to as WPD. While @Gupta2021-hakear already use NQT when computing $\wpd$, it could be useful to standardize it for the selected set of significant granularities prior to computing the distances. In both the JS-NQT and JS-RS approaches, each observation is scaled individually, but in the WPD approach, each observation may or may not be scaled, resulting in usable groups in either scenario.

@@

All of the approaches were fitted to each data design and to each combination of the considered parameters considered in Table \ref{tab:range-parameter}. The formed clusters have to match the design, be well separated, and have minimal intra-cluster variation. <!--It is possible to study these desired clustering traits visually in a more comprehensive way than just looking at index values.--> MDS and parallel coordinate graphs are used to demonstrate the findings, as well as an index value plot to provide direction on the number of clusters. <!--JS-based approaches corresponding to NQT and RS are referred to as JS-NQT and JS-RS respectively.--> In the following plots, results for JS-NQT are reported, and results with JS-RS or wpd-based distances are in the supplementary material.
